Towards Practical First-Order Model Counting

### 2024-04-15 doc/paper.tex
Syntax improvements (and other minor changes)

\begin{abstract}
  TODO: 9 pages!!!
\end{abstract}

@@

\todo[inline,caption={}]{
  Add some papers mentioned in:
  \begin{itemize}
    \item very recent work
    \item my previous paper, including:
    \begin{itemize}
      \item other liftable fragments
      \item some more theory papers, e.g., LICS 2018
    \end{itemize}
  \end{itemize}
}

\paragraph{Papers to cite.}
\begin{itemize}
  \item overviews
        \begin{itemize}
          \item lifted probabilistic inference~\cite{DBLP:conf/ecai/Kersting12}
          \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
        \end{itemize}
  \item Alternative definition~\cite{DBLP:journals/cacm/GogateD16}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item algorithms
        \begin{itemize}
          \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
          \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
          \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
          \item \textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} (similarly to us
                compiles to C++ code, but (probably) doesn't work on as many
                formulas)
          \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
          \item for Markov logic networks~\cite{DBLP:journals/ml/RichardsonD06}
                \begin{itemize}
                  \item \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}
                  \item \textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}
                  \item \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16}
                        (same as the alternative definition)
                \end{itemize}
        \end{itemize}
  \item complexity
        \begin{itemize}
          \item liftability~\cite{DBLP:conf/starai/JaegerB12}
          \item hardness for three variables~\cite{DBLP:conf/pods/BeameBGS15}
          \item liftable fragments
                \begin{itemize}
                  \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
                  \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
                  \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
                  \item some liftable fragments~\cite{DBLP:conf/nips/KazemiKBP16}
                \end{itemize}
        \end{itemize}
  \item applications
        \begin{itemize}
          \item extensions to
                sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
          \item discovery of combinatorial
                sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
          \item conjecturing recurrence
                relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
          \item probabilistic logic
                programming~\cite{DBLP:journals/ijar/RiguzziBZCL17} (WFOMC was
                shown to be supreme)
          \item probabilistic databases~\cite{DBLP:journals/debu/GribkoffSB14}
        \end{itemize}
  \item lifted inference in other domains
        \begin{itemize}
          \item constraint satisfaction~\cite{DBLP:journals/jair/TotisDRK23}
          \item answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}
        \end{itemize}
\end{itemize}

\todo[inline,caption={}]{
  Where to put this?
  \paragraph{Changes to the compilation process.}
  \begin{itemize}
    \item We make the FCGs satisfy the following property. For every domain
          $\Delta$, for every directed path $P$ without Ref nodes, the number of
          atom counting or (generalised) domain recursion or independent partial
          grounding operations on $\Delta$ on $P$ is at most one.
    \item When checking whether recursion is possible between a source and a
          target formula, the target formula (maybe the source formula as well?)
          must not include a domain as well as its subdomain. This might be
          equivalent to the formula being `unshattered'. Such formulas could be
          seen as transitory, and allowing such recursive calls leads to FCGs
          that I cannot make sense of.
  \end{itemize}
}

@@

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item Some experiments vs FastWFOMC
\end{itemize}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item What notation am I using for FOL? Introduce equality cardinality
          constraints, counting quantifiers, etc.
    \item Maybe say ``structure'' instead of interpretation?
    \item Should I say ``sort'' instead of domain everywhere?
  \end{itemize}
}

\begin{table*}
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atomic constraints\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $s = t$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|p| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item allowed quantifiers, and
      \item atomic constructions in addition to those of the form $p(t_{1}, \dots, t_{n})$ for some predicate symbol $p$ and terms $t_{1}, \dots, t_{n}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, the latter of which can depend on the domain size,
      \item $p$ is a predicate symbol, and
      \item $s$ and $t$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

\paragraph{Three types of logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

\todo[inline]{Explain arity and its notation. Define `ground'?}

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{. }\forall x_{2} \in \Delta_{2}\dots\text{
}\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$, where
$\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $p(t_{1}, \dots, t_{m})$, where $p$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant. Throughout the paper, we use set-theoretic notation,
interpreting a formula as a set of clauses and a clause as a set of literals.

For any clause or literal $l$, let $\Vars(l)$ denote the set of variables in
$l$. For any (bound) variable $v$, let $\Dom(v)$ denote the domain over which
$v$ is quantified. For a literal $l$,
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$. Similarly, for any
clause $C$, let $\Doms(C) = \{\, \Dom(v) \mid v \in \Vars(C) \,\}$.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For every predicate $p/n$ in $\phi$, let
  ${(\Delta_{i}^{p})}_{i=1}^{n}$ be a list of the corresponding domains (not
  necessarily distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets such that:
  \begin{itemize}
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{itemize}
  Then an \emph{interpretation} of $\phi$ (with respect to $\sigma$) is a set
  $M$ of ground literals defined by adding either $p(\mathbf{t})$ or
  $\neg p(\mathbf{t})$ for every predicate $p/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{p})$. An interpretation is
  a \emph{model} if it satisfies $\phi$ (see Appendix~A of~Dilkas and
  Belle~\shortcite{DBLP:journals/corr/abs-2306-04189} for more details).
\end{definition}

@@

\paragraph{Notation.}
We write $\expr{}$ for an arbitrary algebraic expression. For any signature or
clause $C$, argument or variable $x$, and number or constant $t$, we shall write
$C[t / x]$ for the result of substituting $t$ for all occurrences of $x$ in $C$.

\begin{itemize}
  \item Variables in uppercase
  \item Domains in uppercase Greek
  \item Constant and predicate symbols in lowercase
  \item Clauses written on separate lines are implicitly conjoined
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item check if I'm still using this notation in all these cases
    \item explain how this works for function calls too
  \end{itemize}
}

\begin{definition}
  A \emph{function call} is a term of the form
  $f(x_{1} - c_{1}, \dots, x_{n} - c_{n})$ (written $f(\mathbf{x} - \mathbf{c})$
  for short), where $f$ is an $n$-ary function, each $x_{i}$ is a variable, and
  each $c_{i}$ is a non-negative constant.
\end{definition}
\todo[inline]{do I ever use this term for the case when some of the variables
  don't exist?}

\begin{definition}
  A \emph{signature} is a term of the form $f(x_{1}, \dots, x_{n})$ (written
  $f(\mathbf{x})$ for short), where $f$ is an $n$-ary function, and each $x_{i}$
  is a variable. The signature of a function call $f(\mathbf{x} - \mathbf{c})$
  is $f(\mathbf{x})$. For example, the signature of $f(x - 1, y - 2)$ is
  $f(x, y)$.
\end{definition}

\begin{definition}
  An \emph{equation} is always of the form $f(\mathbf{x}) = \expr{}$, where
  $f(\mathbf{x})$ is a signature, and $\expr{}$ is an algebraic expression.
  Henceforth, we call $f(\mathbf{x})$ and $\expr{}$ the left-hand side (LHS) and
  the right-hand side (RHS) of the equation, respectively.
\end{definition}
\todo[inline,caption={}]{
  give some examples of formulas that:
  \begin{itemize}
    \item introduce new variables,
    \item call the function itself,
    \item call some other function, and
    \item is more nested than just a sum.
  \end{itemize}
}

@@

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in \mathcal{E}$}{
    \ForEach{function call $f(\mathbf{x} - \mathbf{c}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$c_{i} \in \mathbf{c}$}{
        \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i}/n] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}

@@

\begin{itemize}
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y/0] = f(0, z)$ to $\mathcal{B}$.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
      }
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

@@

\paragraph{Key point.}
If there is a predicate $p$ that has nothing to do with domain $\Delta$, make
sure that $p$ remains as part of the formula even if all clauses with $p$ can be
removed. Which literal is picked on \cref{line:smoothing1} of \cref{alg2} is not
important because any choice achieves the same goal: constructing a clause that
mentions the same predicates as $C'$ and is satisfied by any interpretation.

\begin{fact}
  Assuming that domain $\Delta$ is empty, any clause that contains
  `$\forall x \in \Delta$' (for any variable $x$) is vacuously satisfied by all
  interpretations.
\end{fact}

For example, consider the formula
\begin{align}
  \forall x \in \Delta \forall y, z \in \Gamma &: P(x) \lor Q(y, z)\label[formula]{eq:example1}\\
  \forall y, z \in \Gamma' &: Q(y, z)\label[formula]{eq:example2}
\end{align}
and assume that $\Gamma' \subseteq \Gamma$. In this case, if we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[formula]{eq:simplified}
  \forall y, z \in \Gamma' : Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

We address this issue by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert
\cref{eq:example1,eq:example2} to
\begin{align*}
  \forall y, z \in \Gamma &: Q(y, z) \lor \neg Q(y, z) \\
  \forall y, z \in \Gamma' &: Q(y, z).
\end{align*}
The model count returned by this will also consider the truth value of $Q$ over
$y \in \Gamma \setminus \Gamma'$ or $z \in \Gamma \setminus \Gamma'$.

@@

>>>>

\begin{abstract}
\end{abstract}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages!!!
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
  \end{itemize}
}

\paragraph{Papers to cite.}
\begin{itemize}
  \item overviews
        \begin{itemize}
          \item lifted probabilistic inference~\cite{DBLP:conf/ecai/Kersting12}
          \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
        \end{itemize}
  \item Alternative definition~\cite{DBLP:journals/cacm/GogateD16}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item algorithms
        \begin{itemize}
          \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
          \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
          \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
          \item \textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} (similarly to us
                compiles to C++ code, but (probably) doesn't work on as many
                formulas)
          \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
          \item for Markov logic networks~\cite{DBLP:journals/ml/RichardsonD06}
                \begin{itemize}
                  \item \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}
                  \item \textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}
                  \item \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16}
                        (same as the alternative definition)
                \end{itemize}
        \end{itemize}
  \item complexity
        \begin{itemize}
          \item liftability~\cite{DBLP:conf/starai/JaegerB12}
          \item hardness for three variables~\cite{DBLP:conf/pods/BeameBGS15}
          \item liftable fragments
                \begin{itemize}
                  \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
                  \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
                  \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
                  \item some liftable fragments~\cite{DBLP:conf/nips/KazemiKBP16}
                \end{itemize}
        \end{itemize}
  \item applications
        \begin{itemize}
          \item extensions to
                sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
          \item discovery of combinatorial
                sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
          \item conjecturing recurrence
                relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
          \item probabilistic logic
                programming~\cite{DBLP:journals/ijar/RiguzziBZCL17} (WFOMC was
                shown to be supreme)
          \item probabilistic databases~\cite{DBLP:journals/debu/GribkoffSB14}
        \end{itemize}
  \item lifted inference elsewhere
        \begin{itemize}
          \item constraint satisfaction~\cite{DBLP:journals/jair/TotisDRK23}
          \item answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}
        \end{itemize}
\end{itemize}

@@

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item Completing the definitions of recursive functions by:
  \begin{itemize}
    \item identifying a sufficient set of base cases (\cref{sec:identifying})
    \item constructing formulas that correspond to these base cases
          (\cref{sec:simplifying})
    \item and recursing on these subproblems
  \end{itemize}
  \item Compiling these function definitions into a C++ program that can be
        executed independently for any domain size values \cref{sec:cpp}
        \begin{itemize}
          \item including support for infinite precision arithmetic via GNU
                Multiple Precision Arithmetic Library
        \end{itemize}
  \item Experiments comparing \Cranetwo{} with the main alternative approach
        demonstrate the ability of \Cranetwo{} to scale to domain sizes\dots
\end{itemize}

@@

\begin{table*}
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item allowed quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for some predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k, m \in \mathbb{N}_{0}$, the latter of which can depend on the domain size,
      \item $P$ is a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

\paragraph{Three types of logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item In many-sorted logic, each term is assigned to a \emph{sort}, and each
        predicate $p/n$ is assigned to a sequence of $n$ sorts. Each sort has
        its corresponding domain. In the input formula, all domains are assumed
        to be pairwise disjoin.t Most of these assignments are typically left
        implicit and can be reconstructed from the quantifiers. For instance,
        $\forall x,y \in \Delta\text{. }P(x, y)$ implies that variables $x$ and
        $y$ have the same sort. On the other hand, $\forall x \in \Delta\text{.
        }\forall y \in \Gamma\text{. } P(x, y)$ implies that $x$ and $y$ have
        different sorts, and it would be improper to have $x = y$ as part of a
        formula.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item Add examples of cardinality constraints and counting quantifiers
    \item Mention the role of Skolemization (with a citation)
  \end{itemize}
}

\paragraph{The \FO{} logic (after Skolemization).}
\begin{itemize}
  \item In the spirit of keeping sorts implicit, we always assume formulas `type
        check' with respect to sorts. For example, if $P(x)$, $P(y)$, and
        $x \ne c$ are all part of the formula (for some predicate $P$, variables
        $x$ and $y$, and constant $c$), then $x$, $y$, and $c$ all have the same
        sort.
  \item A \emph{formula} is a conjunction of clauses.
  \item A \emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
        }\forall x_{2} \in \Delta_{2}\dots\text{
        }\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$,
        where $\phi$ is a disjunction of literals that only contain variables
        $x_{1}, \dots, x_{n}$ (and any constants).
  \item A \emph{literal} is either an atom or its negation.
  \item An \emph{atom} is either:
  \begin{itemize}
    \item $P(t_{1}, \dots, t_{m})$ for some predicate $P/m$ and terms
          $t_{1}, \dots, t_{m}$ or
    \item $x=y$ for some terms $x$ and $y$
  \end{itemize}
  \item An atom is \emph{ground} if it contains no variables (i.e., only
        constants).
  \item The \emph{arity} of a predicate is the number of arguments it takes,
        i.e., $m$ in the case of predicate $P$.
  \item When we want to denote a predicate together with its arity, we write
        $P/m$.
  \item A \emph{term} is either a variable or a constant.
  \item Throughout the paper, we use set-theoretic notation, interpreting a
        formula as a set of clauses and a clause as a set of literals. Moreover,
        for readability, clauses written on separate lines are implicitly
        conjoined.
\end{itemize}

For any variable $v$, let $\Dom(v)$ denote the domain over which $v$ is
quantified. For any clause or literal $l$, let $\Vars(l)$ denote the set of
variables in $l$, and, similarly, let
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For every predicate $p/n$ in $\phi$, let
  ${(\Delta_{i}^{p})}_{i=1}^{n}$ be a list of the corresponding domains (not
  necessarily distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets such that:
  \begin{itemize}
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{itemize}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $p(\mathbf{t})$ or
  $\neg p(\mathbf{t})$ for every predicate $p/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{p})$. A structure is a
  \emph{model} if it satisfies $\phi$ (see Appendix~A of~Dilkas and
  Belle~\shortcite{DBLP:journals/corr/abs-2306-04189} for more details).
\end{definition}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item check if I'm still using this notation in all these cases
    \item explain how this works for function calls too
    \item introduce the Iverson bracket (somewhere)
    \item reorder the Logic subsection to cover \FO before comparing it to other
          logics (and clarify the differences between the input format and the
          internal format)
  \end{itemize}
}

\begin{definition}
  A \emph{function call} is $f(x_{1} - c_{1}, \dots, x_{n} - c_{n})$ (written
  $f(\mathbf{x} - \mathbf{c})$ for short), where $f$ is an $n$-ary function,
  each $x_{i}$ is a variable, and each $c_{i}$ is a non-negative constant.
\end{definition}
\todo[inline]{do I ever use this notation for the case when some of the
  variables don't exist?}

\begin{definition}
  A \emph{signature} is $f(x_{1}, \dots, x_{n})$ (written $f(\mathbf{x})$ for
  short), where $f$ is an $n$-ary function, and each $x_{i}$ is a variable. The
  signature of a function call $f(\mathbf{x} - \mathbf{c})$ is $f(\mathbf{x})$.
  For example, the signature of $f(x - 1, y - 2)$ is $f(x, y)$.
\end{definition}

\begin{definition}
  An \emph{equation} is always of the form $f(\mathbf{x}) = \expr{}$, where
  $f(\mathbf{x})$ is a signature, and $\expr{}$ is an algebraic expression.
  Henceforth, we call $f(\mathbf{x})$ and $\expr{}$ the left-hand side (LHS) and
  the right-hand side (RHS) of the equation, respectively.
\end{definition}

\todo[inline,caption={}]{
  an example of a formula that:
  \begin{itemize}
    \item introduces a new variable,
    \item calls itself.
  \end{itemize}
}

@@

\paragraph{Notation.}
We write $\expr{}$ for an arbitrary algebraic expression. In the context of both
algebra and logic, we write $C[x \mapsto y]$ for $C$ with all occurrences of $x$
replaced with $y$.

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{equation $(f(\mathbf{x}) = \expr{}) \in \mathcal{E}$}{
    \ForEach{function call $f(\mathbf{x} - \mathbf{c}) \in \expr{}$}{\label{line:functioncall}
      \ForEach{$c_{i} \in \mathbf{c}$}{
        \For{$n \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto n] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}

@@

\begin{itemize}
  \item \cref{line:insert} for $f(\mathbf{x}) = f(y, z)$, $x_{i} = y$, and
        $n = 0$, add $f(y, z)[y \mapsto 0] = f(0, z)$ to $\mathcal{B}$.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
      }
    }
  }
  \Else{
    $c \gets$ a new constant in $\Delta$\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[v \mapsto c]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

@@

\paragraph{Key point.}
If there is a predicate $P$ that has nothing to do with domain $\Delta$, make
sure that $P$ remains as part of the formula even if all clauses with $P$ can be
removed. Which literal is picked on \cref{line:smoothing1} of \cref{alg2} is not
important because any choice achieves the same goal: constructing a clause that
mentions the same predicates as $C'$ and is satisfied by any structure.

\begin{fact}
  Assuming that domain $\Delta$ is empty, any clause that contains
  `$\forall x \in \Delta$' (for any variable $x$) is vacuously satisfied by all
  structures.
\end{fact}

For example, consider the formula
\begin{gather}
  \forall x \in \Delta\text{. }\forall y, z \in \Gamma\text{. }P(x) \lor Q(y, z)\label[formula]{eq:example1}\\
  \forall y, z \in \Gamma'\text{. }Q(y, z)\label[formula]{eq:example2}
\end{gather}
and assume that $\Gamma' \subseteq \Gamma$. If we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[formula]{eq:simplified}
  \forall y, z \in \Gamma'\text{. }Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

We address this issue by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert
\cref{eq:example1,eq:example2} to
\begin{gather*}
  \forall y, z \in \Gamma\text{. }Q(y, z) \lor \neg Q(y, z)\\
  \forall y, z \in \Gamma'\text{. }Q(y, z).
\end{gather*}
The model count returned by this will also consider the truth value of $Q$ over
$y \in \Gamma \setminus \Gamma'$ or $z \in \Gamma \setminus \Gamma'$.

@@

\todo[inline]{maybe examine \textsc{ForcLift}'s scalability as well}
